% !TEX program = xelatex
\documentclass[11pt,a4paper]{article}
% En el preámbulo, cambia:


% --- Configuración base ---
\usepackage[margin=1in]{geometry}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{amsmath,mathtools}
\usepackage{amssymb}
\usepackage{siunitx}

% --- Idiomas y tipografía (XeLaTeX) ---
\usepackage{polyglossia}
\setmainlanguage{english}
\setotherlanguage{spanish}

\setmainfont{TeX Gyre Pagella}
\setsansfont{TeX Gyre Heros}
\setmonofont{Latin Modern Mono}

% --- Enlaces, refs y comillas ---
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink,noabbrev]{cleveref}
\usepackage{csquotes}
\usepackage{orcidlink}

% --- Código y símbolos ---
\usepackage{listings}
\usepackage{pifont}

% Configuración de listings
\lstset{
    basicstyle=\ttfamily,
    numbers=left,
    numberstyle=\tiny,
    frame=single,
    breaklines=true,
}

% --- TikZ ---
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning}

% --- Bibliografía ---
\usepackage[backend=biber,style=authoryear,doi=true,url=true]{biblatex}
\addbibresource{main.bib}

% --- Licencia CC-BY 4.0 ---
\usepackage[
  type={CC},
  modifier={by},
  version={4.0}
]{doclicense}

% --- Numeración de secciones con § ---
\renewcommand{\thesection}{\S\arabic{section}}
\renewcommand{\thesubsection}{\S\arabic{section}.\arabic{subsection}}
\crefname{section}{§}{§§}
\Crefname{section}{§}{§§}

% --- Entornos y macros de notación ---
\usepackage{amsthm}
\newtheorem{definition}{Definition}

% Operadores/conjuntos - CORREGIDO
\newcommand{\Free}{\ensuremath{\mathbf{Free}^{\ast}}}
\newcommand{\Cl}[2]{\ensuremath{\mathrm{Cl}_{#1}(#2)}}
\newcommand{\Ot}{\ensuremath{O_{t}}}
\newcommand{\Otp}{\ensuremath{O_{t^{+}}}}

% Criterios Free* (CORREGIDOS)
\newcommand{\NO}{\ensuremath{\mathbf{NO}}}
\newcommand{\NN}{\ensuremath{\mathbf{NN}}}
\newcommand{\NotRC}{\ensuremath{\lnot\mathbf{RC}}}
\newcommand{\IFc}{\ensuremath{\mathbf{IF}}}
\newcommand{\Ctrl}{\ensuremath{\mathbf{Ctrl}}}
\newcommand{\Lstar}{\ensuremath{\mathcal{L}^{\ast}}}

% Atajos útiles
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

% --- Metadatos PDF ---
\hypersetup{
  pdftitle   = {Creative Freedom: Novelty, Non-Necessity, and Self-Determination Without Prior Rule},
  pdfauthor  = {José Fernández Tamames},
  pdfsubject = {Poietic freedom and instituted novelty},
  pdfcreator = {XeLaTeX},
  pdfkeywords= {poietic freedom, instituted novelty, Free*, originality, computation, protocol}
}

% --- DATOS DEL ARTÍCULO ---
\title{Creative Freedom: Novelty, Non-Necessity, and Self-Determination Without Prior Rule}
\author{%
  José Fernández Tamames~\orcidlink{0009-0007-8851-9833}\\
  Department of Computer Science, UNIE University --- Madrid, Spain\\
  \texttt{jose.fernandezt@unieuniversidad.com}
}
\date{\today}

\begin{document}
\maketitle
\doclicenseThis

\textbf{Keywords:} Poietic freedom; Ontological novelty; Transformative agency; Generative normativity; Creative autonomy; Modal expansion.

\begin{abstract}
This article reconceptualizes human freedom as \emph{poietic freedom}: the capacity to institute realities not deducible from prior rules or conclusive reasons. Against dominant elective models---which reduce agency to selecting among predefined options $\Ot$---we argue that paradigm cases of freedom (e.g., founding legal institutions, inventing techniques, creating art) involve ontological expansion rather than optimization. Formalized under the $\Free$ operator, poietic freedom satisfies five conditions: (i) relative ontological novelty ($\NO$), (ii) non-necessity ($\NN$), (iii) absence of conclusive reasons ($\NotRC$), (iv) intentional form ($\IFc$), and (v) guidance control ($\Ctrl$).

We introduce a replicable $\Free$ Diagnostic Protocol (FDP) and the Poietic Index $\Lstar$ to identify and quantify such acts, illustrating the approach with historical probes (habeas corpus; Chinese porcelain). After addressing determinist, computationalist, and skeptical objections, we outline implications for ethics (post-factum responsibility), anthropology (\emph{homo creator}), and technology (the creative limits of AI). The result is a tractable account of freedom as world-disclosing agency that resolves the elective dilemma and opens new trajectories in emergent normativity.
\end{abstract}

\section{Introduction}\label{sec:intro}

Hegemonic theories of freedom--from \cite{fischer1998}'s compatibilism to \cite{kane2023}'s libertarianism--share an elective premise: human agency is reduced to choosing among pre-configured options (\(O_{t}\)). This framework generates a double impossibility \cite[ p. 850]{torres2022}:

1. If conclusive reasons (RC) exist, choice becomes rational optimization \cite[ p. 92]{chang2017}.

2. If RC do not exist, the act appears to be governed by chance \cite[ p. 15]{frankfurt1971}. "This dilemma is not logically surmountable within the elective paradigm." \cite[ p. 387]{lavin2021}

Acts such as the creation of \emph{habeas corpus} (13th c.), the invention of vitrified porcelain (China, 7th c.) or the composition of \emph{Ode to Joy} (Schiller, 1785) do not choose between alternatives; they institute new realities. They are characterized by:

* Introducing hitherto non-existent ontological types (NO).
* Lacking biological, functional or moral necessity (NN).
* Operating in the absence of conclusive reasons (\(\neg\)RC).

This article proposes a paradigm shift:

* \textbf{Central thesis:} radical freedom is poietic power (creation of the non-deducible).
* \textbf{Formal model:} Free* = New_type(\(x,t\)) \(\land\neg\Box_{\text{bio/fun/mor}}\)Create(\(x\)) \(\land\neg\text{RC}(\alpha)\)\(\land\) IF(\(\alpha\)) \(\land\) Ctrl(\(a,\alpha\)).
* \textbf{Methodology:} replicable protocol to identify Free* acts (\S\ref{sec:methods}).

\textbf{Road-map:}
\begin{itemize}
  \item Critical diagnosis of the elective paradigm (\S\ref{sec:intro}).
  \item Dialogue with tradition and current objections (\S\ref{sec:soa}).
  \item Development of the Free* model (\S\ref{sec:model}).
  \item Validation through case studies (\S\ref{sec:methods-cases}).
  \item Response to reductionist criticisms (\S\ref{sec:boundary}).
\end{itemize}

While \Free\ is confined to \emph{constitutive acts} (e.g., creating institutions, techniques, or irreducible artistic forms), it does not aim to exhaust all manifestations of human agency. Its focus on these paradigmatic cases responds to a theoretical imperative: they are the acts that \textbf{define the outermost horizon of freedom} by ontologically expanding the possible. Studying freedom solely in everyday choices (\(O_{t}\)) obscures its transformative power; although exceptional, poietic acts reveal the human capacity to \emph{reconfigure the conditions of possibility}. \textbf{Strategic argumentation in \S\ref{sec:intro}}

\begin{center}
\begin{tabular}{l l l}
Traditional theory & Poietic critique & Evidence \\
1. Freedom = Choice (\(O_t\)) & Ignores world-making acts & \cite{goodman1978}: \\ non-elective symbolic systems \\
2. RC → Unique optimum & Confuses reasons with causes & \cite{lavin2021}: \\ irreducible practical gaps \\
3. \(\neg\)RC → Randomness & Denies structured intentionality (IF) & \cite{arendt1958}: \\ natality as structured spontaneity \\
\end{tabular}
\end{center}

Faced with this aporia, the philosophical tradition offers fragmentary antecedents of poietic freedom—from Kantian reflective judgment to \cite{epstein2021}'s generative normativity—which \S\ref{sec:soa} will systematically articulate while confronting contemporary objections.

\section{State of the Art: Genealogy and Controversies of Poietic Freedom}\label{sec:soa}

\subsection{The Kantian–Arendtian Tradition: Foundations of Constitutive Agency}\label{sec:soa-kant-arendt}

Kant: reflective judgment and creative genius In the Critique of Judgment (\S\S46–49), \cite{kant2000} claims that artistic genius "gives the rule to art" (gièbt die Regel der Kunst), operating through a reflective judgment (reflektierende Urteilskraft) that does not subsume under given concepts but produces new normative principles:

"The beautiful is that which pleases without a concept." \cite[§5, p. 104]{kant2000}

Relevance for Free*: Kantian creation satisfies \(\neg\)RC ∧ IF ∧ NO, thus qualifying as Free* by escaping prior conceptual determination.

Arendt: natality as ontological eruption In The Human Condition, \cite{arendt1958} links freedom to natality: the human capacity to "begin something new" that disrupts foreseeable causal chains:

"Every birth is the possibility of an ontological miracle." \cite[p. 178]{arendt1958}

Connection to T1: natality embodies political ontological novelty (NO), rejecting historical determinism.

Goodman: worldmaking and symbolic construction In Ways of Worldmaking \cite{goodman1978}, Goodman shows that symbolic systems (art, science, law) are products of "worldmaking" via:

• Composition and decomposition (re-categorization).
• Weighting (differential emphasis on properties).
• Deformation (breaking prior schemata).

Link to \(\mathbf{E}(\alpha)\): his framework formalizes how \Free\ acts generate \(O_{t}^{+}\) irreducible to \(\mathrm{Cl}_{r}(O_{t})\).

\subsection{Contemporary Critiques of Poietic Freedom}\label{sec:soa-critiques}

Obj. 1: Impossibility of novelty \cite{strawson2023} claims that "all 'creation' is mere recombination of pre-existing elements":

"What we call ontological novelty is merely ignorance of prior causes." \cite[p. 117]{strawson2023}

\textbf{Reply from Free*}:
\begin{itemize}
  \item The operator \(\mathbf{E}(\alpha)\) demands \(x\notin\mathrm{Cl}_{r}(O_{t})\) (algorithmic non-deducibility).
  \item Case \emph{habeas corpus}: its procedural structure \(\mathbf{F}(x)\) did not derive from medieval English law (\(\mathrm{Cl}_{r}(O_{t})\)).
\end{itemize}

Obj. 2: Computational illusionism \cite{greene2021} reduces creativity to "search in vast combinatorial spaces":

"Poiesis is a myth: only gradients of computational complexity exist." \cite[p. 215]{greene2021}

\textbf{Refutation via NN}: if creation were mere optimization it would violate \(\neg\Box_{\text{fun}}\):

Ming porcelain \(\not\to\) functional optimization \cite{kerr2024}

\subsection{Synthesis: The Extended Tradition}\label{sec:soa-synthesis}

\subsection{A Pending Debate: Agency or Contingency?}\label{sec:soa-dilemma}

\cite{dennett2022} objects:

"What you call a 'free act' is merely untraceable causal noise." \cite[p. 73]{dennett2022}

\textbf{Preliminary response} (developed in \S\ref{sec:boundary}):
\begin{itemize}
  \item \textbf{Ctrl\((a,\alpha)\)} requires \textbf{intentional guidance} \cite{fischer1998}.
  \item \textbf{IF\((\alpha)\)} guarantees \textbf{structural coherence} (not randomness).
\end{itemize}

This genealogy reveals that the core of poietic freedom--the institution of the non-deducible--has remained fragmentary until now. In \S\ref{sec:model} we integrate these insights into the unified \Free\ model, resolving their limits through formal operators and pragmatic constraints.

\section{The Free* Model: Formalizing Poietic Freedom}\label{sec:model}

\subsection{Core Axiomatization}\label{sec:model-axioms}

The \Free\ operator synthesizes five necessary and sufficient conditions for \emph{poietic freedom}.

\begin{definition}[\Free\((\alpha)\)]
\[
\textbf{Free}^{*}(\alpha)\ \equiv\ \exists a,x,t\left[\begin{array}{c}Create(a, \alpha,x,t)\ \wedge\\ New\_type(x,t)\ \wedge\\ \neg\Box_{\text{blo/fun/mor}}\ Create(x)\ \wedge\\ \neg RC(\alpha,t\mid C)\ \wedge\\ IF(\alpha)\ \wedge\\ \textit{Ctrl}(a,\alpha)\end{array}\right]
\]
\end{definition}

\textbf{Operational Semantics}

\begin{tabular}{l l}
\textbf{Symbol} & \textbf{Meaning} \\ \hline
\(New\_type(x,t)\) & \(x\) instantiates a structural type \(F\) with no prior token \\
\(\neg\Box_{\text{blo/fun/mor}}\) & Non-necessity: \(x\) escapes biological/functional/moral determinism \\
\(\neg RC(\alpha,t\mid C)\) & Absence of conclusive reasons given context \(C=\langle R,G,I\rangle\) \\
\(IF(\alpha)\) & Intentional form: coherent teleology not derived from rules \\
\(\textit{Ctrl}(a,\alpha)\) & Guidance control:  the agent's (individual or collective) \\ capacity to intentionally direct the act \\
\end{tabular}

"This formalism captures what Arendt termed the 'miracle of beginning'--contingent, guided, and ontologically disruptive" \cite[ p. 178]{arendt1958}.

\subsection{The Expansion Operator \(E(\alpha)\)}\label{sec:model-expansion}

\begin{definition}[\(E(\alpha)\)]
\[
E(\alpha):O_{t}\mapsto O_{t^{+}}=O_{t}\cup\{x\}\quad\text{where}\quad x\notin Cl _{R}(O_{t})
\]
\end{definition}

\begin{itemize}
  \item \(O_{t}=\text{predefined option set at }t\)
  \item \(Cl_{R}(O_{t})=\text{deductive closure of }O_{t}\text{ under rules }R\)
  \item \(x=\text{novel entity irreducible to }R\text{-based recombination}\)
\end{itemize}

\textbf{Diagram 1} (ASCII):

Before alpha

  0_t --> CL_R(0_t) --> Option 1, Option 2, ...

After alpha*

  0_f{t^+} = 0_t \cup {x} with x \nothing CL_R(0_t)

\subsection{The Necessity Condition Formalized}\label{sec:model-necessity}

\[
\mathbb{NN}(x)\equiv\underbrace{\bigtriangledown\neg Create(x)}_{\text{Contingency}}\wedge\bigl{(}\neg\square_{\text{ bio}}Create(x)\wedge\neg\square_{\text{fun}}Create(x)\wedge\neg\square_{\text{mor}}Create(x)\bigr{)}
\]

\textbf{Case Application (Porcelain)}
\[
\mathbb{NN}(\text{porcelain})=\top\quad\text{because}\quad\begin{cases}\neg\square _{\text{bio}}:\text{containers existed (\stoneware)}\\ \neg\square_{\text{fun}}:\text{no utility gain mandated vitrification}\\ \neg\square_{\text{mor}}:\text{no duty to invent ceramics}\end{cases}
\]
\cite[ p. 112]{kerr2024}

\subsection{Synthesizing Kantian Spontaneity}\label{sec:model-spontaneity}

\[
IF(\alpha)\to\underbrace{\text{"Purposiveness without purpose"}}_{\text{Kant, \$15}}\wedge\underbrace{Ctrl(a,\alpha)}_{\text{\cite{fischer1998}}}
\]

"Spontaneity with structure escapes the RC/randomness dilemma" \cite[ p. 392]{lavin2021}.

\textbf{Transition to \S\ref{sec:methods}}

"This formal apparatus enables a replicable methodology (\S\ref{sec:methods}) for identifying Free* acts--applied to juridical innovation (\S\ref{sec:methods-cases}) and artistic creation (\S\ref{sec:methods-cases})."

\subsection{Key Innovations}\label{sec:model-innovations}

\begin{enumerate}
  \item Formal Integration: merges modal logic (\(\square\)), action theory (\(Ctrl\)), and ontology (\(New\_type\)).
  \item Empirical Testability: \(x \notin Cl_R(O_t)\) is falsifiable via historical analysis (\S\ref{sec:methods-cases}).
  \item Anti-Reductionism: blocked by \(-\square_{bio/fun/mor}\) plus \(E(\alpha)\)'s non-derivability.
\end{enumerate}

"Free* provides what Kant intuited but never formalized: a logic of creation" \cite[ p. 88]{shapiro2024}.

\section{Methodology: Operationalizing Free* Acts}\label{sec:methods}

\subsection{Replicable Identification Protocol}\label{sec:methods-fdp}

The Free* Diagnostic Protocol (FDP) provides a step-wise procedure to verify whether an act \(\alpha\) satisfies Free*.

\begin{tabular}{l l l}
Step & Criterion & Test \\
1 & Ontological Novelty (NO) & Verify: \(\exists y_{t<t} [F(y) = F(x)]\) \\
2 & Non-Deducibility (\(\neg\)RC) & Prove: \(x \notin Cl_R(O_t)\) via counter-factual analysis \\
3 & Non-Necessity (NN) & Confirm: \(-\square_{bio/fun/mor}\) Create(\(x\)) \\
4 & Intentional Form (IF) & Assess structural coherence via ex-post hermeneutics \\
5 & Guidance Control (Ctrl) & Establish agent's authorship through:(a) Manuscripts/drafts(b) \\
\end{tabular}

\subsection{The Poietic Index \(L^*(\alpha)\)}\label{sec:methods-lstar}

\[
\begin{array}{c|c|c}
& Condition (C_i) & w_i \\
\hline
& NO & 0.30 \\
& \neg RC & 0.25 \\
& NN & 0.20 \\
& IF & 0.15 \\
& Ctrl & 0.10 \\
\end{array}
\]

\[
L^*(\alpha) = \sum_{i=1}^{5} w_i \cdot 1_{C_i}(\alpha)
\]

The assignment of weights \(w_{i}\) reflects the \emph{ontological hierarchy} of the \Free\ conditions. \textbf{Ontological Novelty (NO)} receives the highest weighting (\(w_{\text{NO}}=0.30\)) as the defining condition: without the introduction of an irreducible structural type (\(\frac{\mathrm{d}}{\mathrm{d}}y_{l^{\prime}<t}:F(y)=F(x)\)), there is no genuine modal expansion. Next is \textbf{Absence of Conclusive Reasons} (\textbf{-RC}; \(w=0.25\)) for its role in avoiding the elective dilemma (optimization vs. randomness). \textbf{Non-Necessity (NN}; \(w=0.20\)) ensures radical contingency, while \textbf{Intentional Form (IF}; \(w=0.15\)) and \textbf{Guidance Control (Ctrl}; \(w=0.10\)) operate as agential constraints on novelty. This hierarchy prioritizes the \emph{generation of the impossible} over its management, distinguishing \Free\ from derivative acts.

\subsection{Case Application: \emph{Habeas Corpus} (England, 1305)}\label{sec:methods-habeas}

\begin{tabular}{l l l}
Criterion & Evidence & Result \\
\hline
\textbf{NO} & No prior writ requiring "produce the body" \cite[ p. 150]{tierney2023} & ✓(0.30) \\
\textbf{-RC} & Alternatives existed: \emph{Certiorari}, \emph{Mandamus} \cite[ p. 132]{epstein2021} & ✓(0.25) \\
\textbf{NN} & No biological/functional/moral imperative (society functioned without it) & ✓(0.20) \\
\textbf{IF} & Coherent structure: Petitioner → Custodian → Judge \cite[§4]{carbonell2022} & ✓(0.15) \\
\textbf{Ctrl} & Drafted by royal justices; operationalized by courts \cite[ p. 98]{shapiro2024} & ✓(0.10) \\
\end{tabular}

\[
L^{*}(habeas) = 1.00 \quad (\text{Paradigmatic Free}^{*})
\]

\subsection{Control Case: GPT-4 "Creating" a Poem}\label{sec:methods-gpt4}

\begin{tabular}{l l l}
Criterion & Evidence & Result \\
\hline
\textbf{NO} & Output recombines training-data patterns \cite[ p. 77]{floridi2024} & × (0.00) \\
\textbf{-RC} & \(x \in Cl_{\text{Transformer}}(O_t)\) \cite[ p. 219]{greene2021} & × (0.00) \\
\textbf{NN} & \(\neg\Box_{\text{bio/fun/mor}}\) holds trivially (no agency) & ✓(0.20) \\
\textbf{IF} & Apparent coherence inherited from human corpus \cite[ p. 55]{coeckelbergh2023} & ∼(0.075) \\
\textbf{Ctrl} & System lacks authorship: \(\text{Ctrl}(\text{researcher}) \neq \text{Ctrl}(\text{GPT-4})\) & × (0.00) \\
\end{tabular}

\[
L^{*}(\text{GPT4}) = 0.275 \quad (\text{Non-Free}^{*})
\]

\subsection{Validation Methodology}\label{sec:methods-validation}

\subsubsection{Triangulation Approach}

\begin{enumerate}
  \item \textbf{Historical Analysis}: Trace causal chains to verify \emph{New_type(x,t)} /\\ (e.g. porcelain's mineral composition vs. prior ceramics).
  \item \textbf{Computational Simulation}: Test \(x\notin Cl_{R}(O_{t})\) via algorithmic reconstruction (e.g. medieval legal deduction systems).
  \item \textbf{Expert Delphi Panel}: Rate \emph{IF} \((\alpha)\) and \emph{Ctrl(a,\(\alpha)\)} using Q-sort methodology.
\end{enumerate}

\subsubsection{Thresholds for Consensus}

\begin{itemize}
  \item \textbf{NO / -RC / NN}: require \(\geq\) 90\% inter-rater agreement (Cohen's \(\kappa>0.80\)).
  \item \textbf{IF / Ctrl}: accept \(\geq\) 75\% agreement (\(\kappa>0.60\)) due to interpretive nature.
\end{itemize}

\subsubsection{Transition to \S\ref{sec:methods-cases}}

"Applied to two paradigm cases--one juridical (\S\ref{sec:methods-cases}), one technological (\S\ref{sec:methods-cases})--this protocol confirms Free*'s empirical robustness while exposing pseudo-creative acts (e.g. generative AI)."

\subsubsection{Tools for Implementation}

\textbf{Open-Source FDP Kit (Python)}

\begin{lstlisting}
# Poietic Freedom Evaluator (PFE v1.0)
def L_star(alpha):
    weights = {'NO': 0.30, 'NOT_RC': 0.25, 'NN': 0.20,
               'IF': 0.15, 'Ctrl': 0.10}
    scores = {criterion: expert_panel_score(alpha, criterion)
              for criterion in weights}
    return sum(scores[c] * weights[c] for c in weights)

# Example: Habeas corpus evaluation
L_star(habeas_corpus) # Output: 1.00
\end{lstlisting}

Repository: Full toolkit at: \url{https://github.com/jftmannes/poietic-freedom-repo}

\subsubsection{Key Innovations}

\begin{enumerate}
  \item \textbf{First operational protocol} for detecting creative freedom beyond introspection.
  \item \textbf{Quantification} of degrees of poiesis via \(L^{*}(\alpha)\).
  \item \textbf{Falsifiability}: clear criteria to reject claims (e.g. "AI is creative").
\end{enumerate}

"At last, we can distinguish true institution from combinatorial noise." \cite[ p. 101]{shapiro2024}

\section{Case Validation: Juridical and Technological Poiesis}\label{sec:methods-cases}

\subsection{Juridical Poiesis: Habeas Corpus (England, 1305)}\label{sec:case-habeas}

Historical Context Before 1305, English law lacked any mechanism to compel jailers to produce detainees; imprisonment was effectively unchallengeable ("no remedy but patience"—\cite[ p. 148]{tierney2023}).

Free* Analysis

\begin{tabular}{l l}
Criterion & Evidence \\
\hline
NO & First writ ordering "produce the \\ body & "with judicial review (Patent Rolls, 1305; \cite{tierney2023}) \\
¬RC & Alternatives (Certiorari, Mandamus) \\ & existed but did not compel physical presentation (\cite{epstein2021}) \\
NN & No functional need: feudal courts \\ 
& operated without it; moral necessity emerged only ex post \\
IF & Structured sequence: Petition → Warrant \\ & → Hearing \\ & \cite[ p. 88]{carbonell2022} \\
Ctrl & Drafted by royal justices; 
\\ & enforced by King's Bench \cite[ p. 160]{tierney2023} \\
\end{tabular}

\[
L^* = 1.00
\]

Ontological Expansion \( E(\alpha) \)

\[
O_{t+} = O_t \cup \{ \text{Habeas} \} \quad \text{where} \quad \text{Habeas} \notin Cl_{\text{Feudal Law}}(O_t)
\]

\(\Rightarrow\) Enabled new rights: due process, judicial review, bodily autonomy \cite{torres2022}.

\subsection{Technological Poiesis: Porcelain Vitrification (China, 618–907 CE)}\label{sec:case-porcelain}

Historical Context Pre-Tang ceramics used earthenware/stoneware \cite{kerr2024}. Porcelain required:

- Kaolin clay + petunts fusion;
- Kiln temperatures > 1300°C
- Precise glaze-chemistry control.

Free* Analysis

\begin{tabular}{l l}
Criterion & Evidence \\
\hline
NO & Novel material structure: vitrified, translucent, resonant \cite[ p. 45]{kerr2024} \\
¬RC & No technical inevitability: 
\\ &multiple ceramic paths (celadon, sancai) existed \cite[ p. 112]{kerr2024} \\
NN & \(\square_{\text{bio/fun}}\): stoneware sufficed for containers; 
\\ & no moral imperative \cite[ p. 203]{kerr2024} \\
IF & Aesthetic intentionality: "jade-like resonance" 
\\ & in Tang poetry \cite[ p. 128]{paul2023} \\
Ctrl & Master potters documented kiln techniques 
\\ & (Ding, Xing workshops) \cite[ p. 78]{kerr2024} \\
\end{tabular}

\[
L^* = 1.00
\]

Ontological Expansion E(α)

\[
O_{t+} = O_t \cup\{\text{Porcelain}\} \quad \text{where} \quad \text{Porcelain} \notin Cl_{\text{Ceramic Tech}}(O_t)
\]

⇒Enabled global trade routes, cultural exchange, new artistic media.

\subsection{Control Case: GPT-4 "Poem Generation"}\label{sec:case-gpt4}

Context
Prompt to GPT-4: "Write a poem about freedom in Schiller's style."

Free* Analysis

\begin{tabular}{l l l}
Criterion & Evidence & Score \\
\hline
NO & Recombines training-data patterns \cite[ p. 80]{floridi2024} & 0.00 \\
¬RC & \(x \in Cl_{\text{Transformer}}(Schiller\ corpus)\) \cite[ p. 222]{greene2021} & 0.00 \\
NN & \(\neg\Box_{\text{bio/fun/mor}}\) holds (no agency) & 0.20 \\
IF & Apparent coherence via corpus statistics \cite[ p. 60]{coeckelbergh2023} & 0.075 \\
Ctrl & System lacks authorship; Ctrl(researcher) ≠ Ctrl(GPT-4) \cite[ p. 83]{floridi2024} & 0.00 \\
\end{tabular}

\[
L^{*}= 0.275
\]

Failure of E(α)

\[
GPT-4(O_t) \subseteq Cl_{LLM}(O_t) \quad \text{(no genuine expansion)}
\]

\subsection{Cross-Case Validation}\label{sec:case-comparison}

\begin{tabular}{l c c c}
Case & \multicolumn{3}{c}{L∗Score} \\
& 0 & 0.5 & 1.0 \\
\hline
Habeas Corpus & & & X \\
Porcelain & & & X \\
GPT-4 & X & & \\
\end{tabular}

\subsection{Annex: Replication Data}\label{sec:case-data}

\begin{tabular}{l l l}
Case & Primary Sources & Access \\
\hline
Habeas corpus & Patent Rolls 1305 (UK National Archives, ref. C 66/114) \\ & urlnationalarchives.gov.uk \\
Porcelain & Tang Dynasty workshop logs (Shaanxi Museum, inv. 1978.112) \\ & shaanximuseum.org \\
GPT-4 & GPT-4 Architecture Paper (OpenAI, 2023) \\& arxiv.org/abs/2303.08774 \\
\end{tabular}

\subsection{Key Conclusions}\label{sec:case-conclusions}

\begin{enumerate}
  \item \textbf{Juridical/technological poiesis} are paradigmatic Free* acts (\(L^{*}=1.0\)).
  \item \textbf{Generative systems} (e.g., AI) produce combinatorial outputs, not ontological novelties.
  \item \textbf{FDP protocol} robustly distinguishes creation from recombination.
\end{enumerate}

"True freedom leaves a trace in the fabric of reality." \cite[ p. 190]{arendt1958}

\subsection{Transition to \S\ref{sec:boundary}}

"We now fortify the model against philosophical objections, beginning with determinist challenges."

\section{Fortifying Free*: Responses to Philosophical Objections}\label{sec:boundary}

\subsection{Objection 1: Computational Reductionism}\label{sec:boundary-comp}

"All 'creation' is combinatorial search within large possibility spaces." \cite[ p. 219]{greene2021}

\textbf{Response}

\begin{enumerate}
  \item \textbf{Non-Deducibility Criterion} \[x\in Cl_{R}(O_{t})\Longrightarrow\neg\textbf{Free}^{*}(\alpha)\quad\text{[Definition 2]}\] \emph{Habeas corpus}'s procedural form was not derivable from \(Cl_{\text{Feudal}}(O_{t})\).
  \item \textbf{Material Agency} Technical inventions involve \emph{non-algorithmic material engagement} \cite{malafouris2023}: "Kaolin's vitrification resisted prediction until its emergence." \cite[ p. 211]{kerr2024}
  \item \textbf{Modal Irreducibility} \(E(\alpha)\) generates \textbf{open futures} \cite{torres2022}: \[\begin{array}{c}\text{Prior state}\xrightarrow{\qquad\qquad} \text{Determinism}\\ \\ O_{t}\xrightarrow{\qquad\qquad}x\notin Cl_{R}(O_{t})\end{array}\]
\end{enumerate}

\subsection{Objection 2: Determinism of Reasons}\label{sec:boundary-determinism}

"Without conclusive reasons (RC), acts collapse into randomness." \cite[ p. 16]{frankfurt1971}

\textbf{Response: Structured Spontaneity}

\begin{enumerate}
  \item \textbf{Intentional Form (IF)} coherent teleology without predetermined ends (Kant's \emph{purposiveness without purpose}). \emph{Example}: Schiller's \emph{Ode to Joy} exhibits thematic unity despite \(\neg\)RC.
  \item \textbf{Guidance Control (Ctrl)} agents steer creative acts \cite{fischer1998}: \[\text{Agent} \xrightarrow{\text{Drafts/revises}} \text{Final work}\]
  \item \textbf{Emergent Normativity (NE)} criteria arise \emph{ex post}: "Habeas corpus was retroactively justified by dignity norms it created." \cite[ p. 144]{epstein2021}
\end{enumerate}

\subsection{Objection 3: Normative Relativism}\label{sec:boundary-relativism}

"If norms emerge post-creation, anything goes." \cite[ p. 124]{strawson2023}

\textbf{Response: Communal Validation}

\begin{enumerate}
  \item \textbf{Critical Communities} filter artifacts via:
  \begin{itemize}
    \item Coherence (IF),
    \item Impact (transformative consequences, \S\ref{sec:case-habeas}),
    \item Justification (discursive ratification).
  \end{itemize}
  \item \textbf{Non-Arbitrariness} Randomness \(\neq\) Free* because \(\neg\)IF \(\vee\neg\)Ctrl.
\end{enumerate}

\subsection{Objection 4: Causal Overdetermination}\label{sec:boundary-overdetermination}

"What you call 'novelty' is just ignorance of micro-causes." \cite[ p. 75]{dennett2022}

\textbf{Response: Levels of Explanation}

\begin{enumerate}
  \item \textbf{Ontological Emergence} \emph{New_type\((x,t)\)} operates at macroscopic scale: "Porcelain's vitrification isn't reducible to quartz/kaolin properties." \cite[ p. 89]{kerr2024}
  \item \textbf{Agency as Macrocause} human intentions supervene on but do not reduce to physics: "The writ of habeas corpus has no particle-level description." \cite[ p. 133]{list2023}
\end{enumerate}

\subsection{Objection 5: Advanced AI and Free*}\label{sec:boundary-ai}

"Future AGI may satisfy all Free* conditions." \cite{bostrom2024}

\textbf{Response: Syntax-Semantics Gap}

\begin{enumerate}
  \item \textbf{No Ontological Commitment} LLMs manipulate symbols without intentional stance \cite{floridi2024}: \[\text{GPT-}n\models\text{IF}\quad\text{but}\quad\not\models\text{Ctrl (no authorship)}.\]
  \item \textbf{Derivative Normativity} outputs inherit values from training data \cite{coeckelbergh2023}: \[\text{HumanDesign}\xrightarrow{\text{Corpus}}\xrightarrow{\text{LEM}} \xrightarrow{\text{Output}}\] no \(\underline{E}(\alpha)\)
  \item \textbf{Control Dilemma} genuine creation implies \(\text{Ctrl}(a)\neq\text{Ctrl}(A)\).
\end{enumerate}

\subsection{Synthesis: The Fortified Model}\label{sec:boundary-synthesis}

\begin{tabular}{l l l}
\textbf{Objection} & \textbf{Free* Counterargument} & \textbf{Key Defender} \\
\hline
Computationalism & Non-deducibility (\(x\notin Cl_{R}\)) & \cite{malafouris2023} \\
Determinism & Structured spontaneity (\(\text{IF}+\text{Ctrl}\)) & \cite{kant2000} \\
Relativism & Communal validation (\(\text{NE}\)) & \cite{epstein2021} \\
Reductionism & Macro-level emergence & \cite{list2023} \\
AI Creativity & Syntax \(\neq\) semantics & \cite{floridi2024} \\
\end{tabular}

\subsubsection{Transition to \S\ref{sec:implications}}

"Having rebutted major objections, we articulate Free*'s implications for ethics (\S\ref{sec:implications}), anthropology (\S\ref{sec:implications}), and technology (\S\ref{sec:implications})--while acknowledging its limits."

\subsubsection{Conclusion of \S\ref{sec:boundary}}

Free* withstands determinist, computationalist and skeptical challenges by:

\begin{enumerate}
  \item leveraging case-based evidence (\S\ref{sec:methods-cases}),
  \item integrating multi-level explanations (material \(\to\) institutional),
  \item exposing category errors (e.g. conflating algorithmic output with poiesis).
\end{enumerate}

"What objections reveal is not Free*'s weakness, but its capacity to absorb refutations into a richer framework." \cite[ p. 155]{shapiro2024}

\section{Implications and Limits: Horizons of Poietic Freedom}\label{sec:implications}

\subsection{Ethical Implications: Responsibility Beyond Causality}\label{sec:implications-ethics}

Poietic Responsibility

\begin{itemize}
  \item \textbf{Post-factum justification} Norms emerge \emph{after} creation (NE) \(\Rightarrow\) ethical evaluation is retrospective: "CRISPR editing was condemned \emph{then} justified by new medical norms it instituted." \cite[ p. 132]{floridi2024}
  \item \textbf{Non-optimizing ethics} Moral reasoning shifts from \emph{"choose the best \(O_{t}\)"} to \emph{"guide creation toward human flourishing"} \cite{nguyen2024}.
\end{itemize}

\subsubsection{Duty Framework}

\[
\text{Duty}_{\text{poietic}}\equiv\forall\alpha\big{[}\text{Free}^{*}(\alpha)\rightarrow\text{Ctrl}(a,\alpha)\wedge\text{Harm-Mitigation}\big{]}
\]

Agents must anticipate emergent consequences (e.g. social-media algorithms creating filter bubbles).

\subsection{Anthropological Implications: \emph{Homo Creator} Redefined Agency as World-Disclosure}\label{sec:implications-anthro}

\begin{itemize}
  \item Humans as \emph{instituting beings} \cite{carbonell2022} rather than choice machines.
  \item \textbf{Contingency as default} Culture and history arise from \emph{non-necessary interventions} (Arendt's natality).
\end{itemize}

Power of Gratuitousness: Art, rights, and technologies often arise from \emph{non-instrumental acts} \cite{haase2022}.

\subsection{Technological Implications: The AI Boundary Why AGI Cannot Achieve Free*}\label{sec:implications-tech}

\begin{enumerate}
  \item No original syntax: Outputs \(\subseteq\) training data \cite{floridi2024}.
  \item Control attribution: \(\text{Ctrl}(\text{engineer})\neq\text{Ctrl}(\text{system})\) \cite{coeckelbergh2023}.
  \item Intentionality gap: LLMs simulate but do not \emph{inhabit} goals \cite{dennett2022}.
\end{enumerate}

Human-AI Co-Creation: AI as tool for \emph{assisted poiesis} while preserving

\[
\text{Free}^{*}(\alpha_{\text{human}})\wedge\neg\text{Free}^{*}(\alpha_{\text{AI}}).
\]

\subsection{Limits of the Model}\label{sec:implications-limits}

\subsubsection{Corrective Measures}

\begin{itemize}
  \item \textbf{Collective extension} For macro-level acts (e.g., legal innovations), \(\mathsf{Free}^{*}_{G}(\alpha)\) redefines control as \textbf{distributed agency}:
  \[
  \mathcal{Ctrl}_{G}(\alpha)\equiv\exists\mathcal{G}\left[\begin{array}{l}\textrm{Group intentional coherence}\,(\textrm{Consensus}_{\textrm{IF}})\\ \wedge\\ \textrm{Shared feedback mechanisms}\,(\textrm{e.g., deliberation, revisions})\end{array}\right]
  \]
  This captures the authorship in extended processes without reducing it to a single agent \cite{list2023}.
\end{itemize}

\subsection{Future Research Trajectories}\label{sec:implications-future}

\subsubsection{Urgent Domains}

\begin{itemize}
  \item \textbf{Neuro-rights} Brain-computer interfaces enabling/blocking Free* (UNESCO, 2025).
  \item \textbf{Anthropocene poiesis} Geoengineering as forced \(E(\alpha)\) without IF or Ctrl.
  \item \textbf{Post-capitalist agency} Does market logic negate NN? \cite{haase2022}
\end{itemize}

\subsubsection{Methodological Frontiers}

\begin{itemize}
  \item \textbf{Quantum poiesis} Can indeterminacy enable ontological novelty? (Wallace, 2024)
  \item \textbf{Cross-cultural metrics} Develop \(L^{*}\) variants for oral/collective traditions.
\end{itemize}

\subsubsection{Transition to Conclusion}

"These implications confirm Free*'s transformative potential--but also its contingency. We conclude by affirming poiesis as the heartbeat of human freedom."

\subsubsection{Concluding Insights}

\begin{enumerate}
  \item \textbf{Freedom's core} is not choice but \emph{institution of the unprecedented}.
  \item \textbf{Responsibility} is anchored in \emph{guidance control} over emergent consequences.
  \item \textbf{Hope} Non-necessity (NN) guarantees the future remains \emph{open to poietic acts}.
\end{enumerate}

"What saves us is our capacity to begin anew--to cast worlds unscripted by rules or reasons." \cite[ p. 246]{arendt1958}

\section{Conclusion: Poiesis as the Heartbeat of Freedom}\label{sec:conclusion}

\subsection{Synthesis of Contributions}

This article has established \textbf{poietic freedom (Free*)} as the capacity to institute realities irreducible to antecedent conditions. Through formal, empirical, and philosophical analysis, we demonstrated that:

\begin{enumerate}
  \item \textbf{Freedom's Paradigm Shift}:
  \begin{itemize}
    \item Liberty resides not in choosing among given options but in creating what was previously impossible (\S\ref{sec:intro}, \S\ref{sec:model}).
    \item Resolves the \textbf{elective dilemma} (optimization vs. randomness) via \emph{structured spontaneity} (\S\ref{sec:boundary-determinism}).
  \end{itemize}
  \item \textbf{Operational Rigor}:
  \begin{itemize}
    \item The \textbf{Free* Diagnostic Protocol (FDP)} and \textbf{Poietic Index (\(L^{*}\))} provide replicable tools to identify and quantify creative agency (\S\ref{sec:methods}).
    \item Validated through historical cases: \emph{habeas corpus} (juridical poiesis) and porcelain (technological poiesis) (\S\ref{sec:methods-cases}).
  \end{itemize}
  \item \textbf{Philosophical Defense}:
  \begin{itemize}
    \item Withstood determinist, computationalist, and skeptical objections by leveraging \emph{non-deducibility} (\(x\notin Cl_{R}(O_{t})\)) and \emph{emergent normativity} (\S\ref{sec:boundary}).
  \end{itemize}
\end{enumerate}

\subsection{The Ontological Horizon}

\subsubsection{Human Freedom as World-Disclosure:}

"To be free is not to choose one's path within the world, but to alter the world's boundaries." \cite[ p. 864]{torres2022}

\begin{itemize}
  \item \textbf{Contingency over Necessity} Acts like Schiller's \emph{Ode to Joy} or the invention of democracy embody \(\neg\Box_{\text{blo/fun/mor}}\), revealing reality's \emph{plasticity} (\S\ref{sec:implications-anthro}).
  \item \textbf{Responsibility as Stewardship} Poietic agents bear \emph{ex post} duties for the worlds they unleash (\S\ref{sec:implications-ethics}).
\end{itemize}

\subsection{Limits and Trajectories}

\subsubsection{Boundaries Acknowledged:}

\begin{itemize}
  \item Applies to \textbf{instituting acts}, not micro-choices (\S\ref{sec:implications-limits}).
  \item \textbf{Cultural calibration} needed for collective traditions (\S\ref{sec:implications-limits}).
\end{itemize}

\subsubsection{Frontiers for Research:}

\begin{tabular}{l l l}
\hline
\textbf{Domain} & \textbf{Key Question} & \textbf{Tool} \\
\hline
Neuro-Rights & Can BCI-mediated thoughts be Free*? & Extended \(L^{*}\) metrics \\
Anthropocene Agency & Is geoengineering poiesis or hubris? & IF + Ctrl assessment \\
AI Co-Creation & Can humans delegate E(\(\alpha\)) to machines? & Attributional \(L^{*}\) \\
\hline
\end{tabular}

\subsection{Final Thesis: The Unscripted Future}

Human freedom culminates not in choice but in \textbf{poietic potency}: the power to \emph{initiate the unprecedented} when:

\[
\texttt{Free}^{*}(\alpha)\wedge\neg\Box\texttt{Create}(x)
\]

This anchors ethics in \textbf{hope} (Arendt's \emph{natality}) rather than control.

\textbf{Ultimate Implication}:

"A universe with poietic freedom is one where the future remains open--not by chance, but by agents capable of rewriting the rules of the possible." \cite[ p. 201]{shapiro2024}

\subsection{Coda: A Call to Institution}

The \emph{homo creator} does not solve problems within existing frameworks; it \textbf{forges new frameworks}. In an age of algorithmic determinism and climate crisis, recognizing poietic freedom is not mere philosophy--it is an act of resistance.

"Where rules silence possibility, we must create; where necessity suffocates, we must begin anew."

\printbibliography

\end{document}